---
title: "5_differential_abundance"
author: "Pedro"
date: "9/20/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---
## 5.0 Deseq2 differential abundance analysis on lists of comparisons
The use of lists in this analysis greatly reduces the amount of hard code. if running this for the first time, read carefully about the use of deseq2 for amplicon data. Each dataset may need tweaks on the selected p levels or minimal fold changes; for that you will edit build_sig_tab()


base tutorial used here: https://joey711.github.io/phyloseq-extensions/DESeq2.html
heatmap for final plot: http://rstudio-pubs-static.s3.amazonaws.com/288398_185f2889a5f641c6b9aa7b14fa15b635.html

### 5.1 Load libraries and create deseq2 object
This will be the bulk of the calculations and can take a moment

```{r}
ps_list_nonraref<-phyloseq_sep_variable(physeq_filtered, variable = c("Plant_species","Sample_type")) # this makes one list wth 4 NON-RAREFIED objects


# make list of deseqed phyloseq objects
ps_list_deseq <- lapply (ps_list_nonraref, function (x) phyloseq_to_deseq2(x, ~ MeJA_treatment))
ps_list_deseq <- lapply (ps_list_deseq, function (x) DESeq(x, test="Wald", fitType="parametric"))

# this unlisted, unsplit deseq2 object will be necessary to construct the heatmap
ps_deseq <- phyloseq_to_deseq2(physeq_filtered, ~ MeJA_treatment)
ps_deseq <- DESeq(ps_deseq, test="Wald", fitType="parametric")

# make a list of the pairwise comparisons you want to run according the treatment ; 
#  positive fold change: higher in fist factor (treatment) / negative fold change: higher in second factor(control)
pairwise_comparison<-list(
c("MeJA_treatment", "MeJA_0.1", "control"),
c("MeJA_treatment", "MeJA_1.0", "control"), 
c("MeJA_treatment", "oral_secretion", "control"),
c("MeJA_treatment", "oral_secretion", "MeJA_0.1"),
c("MeJA_treatment", "oral_secretion", "MeJA_1.0"),
c("MeJA_treatment", "MeJA_0.1", "MeJA_1.0"))

```
### 5.2 make the pairwise comparisons
These comparisons will be based on the data splits and the pairwise_comparison list you generated in the previous chunk

```{r}
# this function will build the significance table for each list, and has to be run once for each pairwise comparison (contro_VS_Meja1.0 and Control_VS_MeJA0.1)
# I don't know why it is not working for the last contrast of the list above. it could be due to different sample sizes

build_sig_tab<-function (deseq_list, contrast_cha_vector){

  alpha<-0.05 # defines the alpha level for p adjustment to control for false discovery rate
  min_fold<-4 # defines minimum fold difference in counts to classify an OTU as differential


  list_res <- lapply(deseq_list, function(x) 
    results(x, contrast=contrast_cha_vector, cooksCutoff = FALSE))

  list_sigtab <- lapply(list_res, function (x) 
    x[which(x$padj < alpha), ]) # cuts results based on alpha for FDR
  
   list_sigtab <- lapply(list_sigtab, function (x) 
    x[which(abs(x$log2FoldChange) > min_fold), ]) # cuts results based on minimum fold change
   
  list_sigtab <- lapply(list_sigtab, function (x) 
    x[order(x$padj),])  # sorts lists of OTUs based on p values of pairwise abundance differences
  
  
  
  
  
  
  # this silenced chuck allows you to limit the number of OTUs accepted as diferential in each comparision. note that if you define the minimum number as 1:40 (1 to 40) but you only hace 39 differentially abundannt OTUs the function will crash and you will not have these results. just remove the hashtags below to limt to 3 top differentially abundant OTUs
  
    # max_OTU_n<-1:3 # defines the maximum number of OTUs to include in each pairwise coparison
    # list_sigtab <- lapply(list_sigtab, function (x) 
    # x[max_OTU_n,])  # cuts results based on maximum numer of OTUs counted as differential
  
  
  
  
  
  
    
  list_sigtab <- lapply(list_sigtab, function (x) 
    cbind(as(x, "data.frame"), as(tax_table(physeq_filtered)[rownames(x), ], "matrix"))) # here we only colect the taxa names from the original, unsplit, phyloseq object
  return(list_sigtab)
}





#now use the custom function to build the significance tables for each comparison defined in the contrasts. 
# these represent only treatment vs control comparisons
Sigtab1<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[1]])
Sigtab2<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[2]])
Sigtab3<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[3]]) # we can use lists here too, mayme mapply() or map() but I won't fix it at this point



Sigtab4<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[4]])
Sigtab5<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[5]])
Sigtab6<-build_sig_tab(deseq_list = ps_list_deseq, 
                       contrast_cha_vector = pairwise_comparison[[6]])



#now extract the rownames of the significance tables and put it into a single vector
select_listed <- do.call(c, list(Sigtab1, Sigtab2, Sigtab3,Sigtab4, Sigtab5, Sigtab6))%>% #concatenate the lists resulting from build_sig_tab(), then...
  lapply(function (x) rownames(x))%>% # extract the rownames from the list, then....
unlist(use.names=FALSE)%>% # flatten the rowname (OTU ID) list, ignoring the names of the lists, then...
  unique() # remove duplications of the same OTUs


```
### 5.3 Construct heatmaps
A heatmap like this is a convinient way to represent the key differences across OTUs and treatments

note that on this heatmap, all comparisons are made against the control. you will need to change the contrasts to make, for example, a 1.0 Meja - 0.1MeJA comparison
```{r}
# finally, build the heatmap
# note that the log change calculations are based on the full, unsplit and unlisted deseq2 object (that is, our complete dataset)
ntb <- normTransform(ps_deseq) # defaults to log2(x+1)
log2.norm.countsb <- assay(ntb)[select_listed, ] #these colnames should match rownames(df)
df <- as.data.frame(colData(ps_deseq)[,c("MeJA_treatment", "Sample_type", "Plant_species")])
row.names(df)<-row.names(as.data.frame(colData(ps_deseq)))
df_row <- as.data.frame((tax_table(physeq_filtered))[,c("Class")])
pheatmap::pheatmap(log2.norm.countsb, color=viridis(24), annotation_col=df, annotation_row=df_row, main="log2(counts + 1)", 
                   scale="none", fontsize_row = 3, fontsize_col = 8)




# if you are dealing with too many samples and it's difficult to read, plot one list element at a time
ntb <- normTransform(ps_list_deseq[[4]]) # defaults to log2(x+1)
differential_OTUs_in_deseq_list_element <- select_listed[which(select_listed%in%row.names(assay(ntb)))]# subsets selection of differentially abundant OTUs so that only those represented in the deseq2 list portion are present
log2.norm.countsb <- assay(ntb)[differential_OTUs_in_deseq_list_element, ] #these colnames should match rownames(df)
df <- as.data.frame(colData(ps_list_deseq[[4]])[,c("MeJA_treatment", "Sample_type", "Plant_species")])
row.names(df)<-row.names(as.data.frame(colData(ps_list_deseq[[4]])))
df_row <- as.data.frame((tax_table(ps_list[[4]]))[,c("Class")])
pheatmap::pheatmap(log2.norm.countsb, color=viridis(24), annotation_col=df, annotation_row=df_row, main="log2(counts + 1)", 
                   scale="none", fontsize_row = 8, fontsize_col = 8)



# if you are dealing with too many samples and it's difficult to read, plot one list element at a time with this function:
mapply (function(list_deseq_objects, list_ps_objects) {

ntb <- normTransform(list_deseq_objects) # defaults to log2(x+1)
differential_OTUs_in_deseq_list_element <- select_listed[which(select_listed%in%row.names(assay(ntb)))]# subsets selection of differentially abundandt OTUs so that only those represented in the deseq2 list portion are present
log2.norm.countsb <- assay(ntb)[differential_OTUs_in_deseq_list_element, ] #these colnames should match rownames(df)
df <- as.data.frame(colData(list_deseq_objects)[,c("MeJA_treatment", "Sample_type", "Plant_species")])
row.names(df)<-row.names(as.data.frame(colData(list_deseq_objects)))
df_row <- as.data.frame((tax_table(list_ps_objects))[,c("Class")])
pheatmap::pheatmap(log2.norm.countsb, color=viridis(24), annotation_col=df, annotation_row=df_row, main="log2(counts + 1)", 
                   scale="none", fontsize_row = 3, fontsize_col = 8)

},
list_deseq_objects = ps_list_deseq,
  list_ps_objects = ps_list_rarefied,
  SIMPLIFY = FALSE)

















```



### 5.4 plot a logfoldchange summary
this is the base sample from the tutorial

```{r}
library("ggplot2")

theme_set(theme_bw())
scale_fill_discrete <- function(palname = "Set1", ...) {
    scale_fill_brewer(palette = palname, ...)
}

# Phylum order
x = tapply(Sigtab1$Arabidopsis_thaliana.Root$log2FoldChange, Sigtab1$Arabidopsis_thaliana.Root$Phylum, function(x) max(x))
x = sort(x, TRUE)
Sigtab1$Arabidopsis_thaliana.Root$Phylum = factor(as.character(Sigtab1$Arabidopsis_thaliana.Root$Phylum), levels=names(x))
# Genus order
x = tapply(Sigtab1$Arabidopsis_thaliana.Root$log2FoldChange, Sigtab1$Arabidopsis_thaliana.Root$Genus, function(x) max(x))
x = sort(x, TRUE)
Sigtab1$Arabidopsis_thaliana.Root$Genus = factor(as.character(Sigtab1$Arabidopsis_thaliana.Root$Genus), levels=names(x))
ggplot(Sigtab1$Arabidopsis_thaliana.Root, aes(x=Genus, y=log2FoldChange, color=Phylum)) + geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))





lapply()











library("ggplot2")

theme_set(theme_bw())
scale_fill_discrete <- function(palname = "Set1", ...) {
    scale_fill_brewer(palette = palname, ...)
}

# Phylum order
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))
ggplot(sigtab, aes(x=Genus, y=log2FoldChange, color=Phylum)) + geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))









```












If your analyis is difficult to read because it has too many OTUs, you might want to decrease the aplha, inrease the minimum fold or activate the maximimum number of otus in the custom build_sig_tab() function

### 5.4 add differential abudance classification to phyloseq object 
```{r}


########### automate
# i need 4 lists (sp/niche) of OTUs that are differential related to contorl or not
# this classification will be simplified: is it more abundant in any treatment when compared to control?
# this way they can be ploted on the network or other visualizations/tables



# puts all sigtabs from one environment in only 2 strings...
abundant_taxa_list<-mapply (function (x, y,z){
# the input here is the output of the build_sig_tab() custom function  
  
#get the row names, but only of ASVs that satisfity the condition based on size of logfold change
abundand_in_treatment1<-row.names(x[x$log2FoldChange>0,]) #more common in treatment
abundand_in_control1<-row.names(x[x$log2FoldChange<0,]) #more common in control

abundand_in_treatment2<-row.names(y[y$log2FoldChange>0,]) #more common in treatment
abundand_in_control2<-row.names(y[y$log2FoldChange<0,]) #more common in control

abundand_in_treatment3<-row.names(z[z$log2FoldChange>0,]) #more common in treatment
abundand_in_control3<-row.names(z[z$log2FoldChange<0,]) #more common in control

#redeplicate, removing repetitive ASV entries
abundand_in_treatment<-unique(c(abundand_in_treatment1, abundand_in_treatment2,abundand_in_treatment3))
abundand_in_control<-unique(c(abundand_in_control1, abundand_in_control2,abundand_in_control3))

# put it into a list, and  give names to the list portions
output<-list(abundand_in_treatment,abundand_in_control)
names(output)<-c("abundand_in_treatment","abundand_in_control" )

return(output)

},

x = Sigtab1, 
y = Sigtab2,
z = Sigtab3,
SIMPLIFY = FALSE)



# now check if any OTU is above in both control AND one of the treatments (suchas abudance Meja0.1>control>Mejas1.0 )
lapply(abundant_taxa_list, function (x) intersect(x$abundand_in_treatment, x$abundand_in_control)) # do this lists share a value?
lapply(abundant_taxa_list, function (x) (x$abundand_in_treatment%in%x$abundand_in_control)) #is list element 1 found in list element2?
lapply(abundant_taxa_list, function (x) (x$abundand_in_control%in%x$abundand_in_treatment)) #is list element 2 found in list element1?



# this makes a new phyloseq object, adding the deseq2 classifications over a list
list_ps_with_deseq2<-mapply(function (list_ps_object,abundant_taxa_list_output){
  output<-mutate( #make a new df wiht a new column...
  as.data.frame(tax_table(list_ps_object)), # on the dataframe-converted  taxa table of your phyloseq object,
       deseq2_ASV_class = if_else( # the new column is named deseq2_ASV_class...
         taxa_names(list_ps_object)%in%
           abundant_taxa_list_output$abundand_in_treatment == TRUE , "abundant_in_treatment", if_else( #if the taxa name is on the list of taxa classified as abudand in treatment, give this claissification on the enw variable, or else...
             taxa_names(list_ps_object)%in%
              abundant_taxa_list_output$abundand_in_control == TRUE, "abundant_in_control", "non-differential")))# check if the taxa is abundant in control and add the tag. if it is not in the list, tag it as non-differential
     tax_table(list_ps_object)<-merge_phyloseq(list_ps_object, tax_table(as.matrix(output))) #now update the tax_table of the original phyloseq object
     
},
  list_ps_object = ps_list_rarefied,
  abundant_taxa_list_output = abundant_taxa_list,
  SIMPLIFY = FALSE)



tax_table(ps_list_rarefied$Arabidopsis_thaliana.Root)

write.csv(tax_table(list_ps_with_deseq2$Arabidopsis_thaliana.Root)[,9], "AT_root_deseq.csv")
write.csv(tax_table(list_ps_with_deseq2$Arabidopsis_thaliana.Soil)[,9], "AT_soil_deseq.csv")
write.csv(tax_table(list_ps_with_deseq2$Brassica_oleraceae.Root)[,9], "BO_root_deseq.csv")
write.csv(tax_table(list_ps_with_deseq2$Brassica_oleraceae.Soil)[,9], "BO_soil_deseq.csv")




       



```